% Just The Docs Front Matter
% title: Debugging
% parent: Troubleshooting
% nav_order: 6
% has_children: false
% has_toc: false

\section{Debugging} \label{sec:troubleshooting-debugging}
\subsection*{How to debug a crash in issm.exe?}
If there is crash during the solve phase, we strongly suggest using \href{http://valgrind.org}{Valgrind}. Install Valgrind using one of the script in the directory \lstinlinebg|${ISSM_DIR}/externalpackages/valgrind|. Valgrind will subsequently be embedded in ISSM and can detect segmentation faults as well as memory leaks. To do so, set the model debugging field to 1 and use only one CPU,
\begin{lstlisting}
md.debug.valgrind = 1;
md.cluster.np = 1;
\end{lstlisting}
in your script, or, simply set,
\begin{lstlisting}
valgrind = true;
\end{lstlisting}
in \lstinlinebg|src/m/classes/debug.m| to apply run Valgrind on all subsequently-run scripts.
Launch the solution sequence and read the \lstinlinebg|errlog| file that it outputs.

\subsubsection{When a build includes Boost}
If your build includes the Boost C++ libraries, there are additional configuration steps needed to overcome a conflict when running Valgrind. Either,
\begin{enumerate}
\item install Valgrind with \lstinlinebg|externalpackages/valgrind/install-<OS>.sh|
\item set \lstinlinebg|#define BOOST_MATH_PROMOTE_DOUBLE_POLICY false| in \newline\lstinlinebg|externalpackages/valgrind/src/boost/math/tools/user.hpp| before running \lstinlinebg|bootstrap.sh| and \lstinlinebg|b2 install|
\item if you are using Dakota, this will need to be reinstalled as well
\end{enumerate}

\subsection*{Interfaces}
\subsubsection{How to debug a MATLAB crash?}
If there is a crash that is not in \lstinlinebg|issm.exe| (sometimes shown as by PETSc's error manager), one should also use Valgrind. Use the following command,
\begin{lstlisting}
matlab -nojvm -nosplash -r "your matlab commands" -D"valgrind --error-limit=no --tool=memcheck -v --log-file=valgrind.log"
\end{lstlisting}
Valgrind's output file \lstinlinebg|valgrind.log| should help (look for Invalid read and Invalid write).

\subsubsection{How to debug a Python crash?}
If there is a crash that is not in \lstinlinebg|issm.exe| (sometimes shown as by PETSc's error manager), one should also use Valgrind. Use the following command:
\begin{lstlisting}
valgrind  --error-limit=no --tool=memcheck -v --log-file=valgrind.log python -E -tt ./yourpythonscript.py
\end{lstlisting}
Valgrind's output file \lstinlinebg|valgrind.log| should help (look for Invalid read and Invalid write).

\emph{NOTE}: if line numbers are not displayed for Mac users, add the following option \lstinlinebg|--dsymutil=yes|

\clearpage % Make sure all figures are placed before next section
