% Just The Docs Front Matter
% title: Loading ISSM
% parent: Getting Started
% nav_order: 1
% has_children: false
% has_toc: false

\section{Loading ISSM} \label{sec:getting-started-loading-issm}
By default, MATLAB and Python cannot locate ISSM and its external packages. We outline here how to load ISSM in your chosen interface.

\subsection{Environment}
First, we need to make sure our interface can find ISSM and its external packages by running,

\lstinlinebg|bash, zsh|
\begin{lstlisting}
export ISSM_DIR=<ISSM_PATH>
source ${ISSM_DIR}/etc/environment.sh
\end{lstlisting}

\lstinlinebg|csh|
\begin{lstlisting}
setenv ISSM_DIR <ISSM_PATH>
source ${ISSM_DIR}/etc/environment.csh
\end{lstlisting}

where \lstinlinebg|<ISSM_PATH>| is the path to ISSM on disk (ex: \lstinlinebg|${HOME}/ISSM/src|). This works whether you downloaded one of our precompiled distributables or compiled ISSM from source.

\subsection{Scripting Interfaces}

\subsubsection{MATLAB}

% TODO: Explore using startup file to modify environment by source'ing $ISSM_DIR/etc/environment.sh: https://www.mathworks.com/help/matlab/ref/startup.html

%__@LATEX_ONLY_START@__
\emph{NOTE}:
There does not currently exist a method for starting the MATLAB GUI using its launcher icon with ISSM loaded.
%__@LATEX_ONLY_END@__
%__@MARKDOWN_ONLY_START@__
%{: .highlight-title }
%> Note
%>
%> There does not currently exist a method for starting the MATLAB GUI using its launcher icon with ISSM loaded.
%__@MARKDOWN_ONLY_END@__

Start MATLAB in a terminal by running,
\begin{lstlisting}
matlab
\end{lstlisting}
If the \lstinlinebg|matlab| executable is not on your environment's path, you will need to add its parent directory to the \lstinlinebg|PATH| environment variable, create an alias for \lstinlinebg|matlab|, or start MATLAB with the full path to the \lstinlinebg|matlab| executable.

After starting MATLAB, you must tell it where to find ISSM's binaries and libraries by running,
\begin{lstlisting}
addpath <ISSM_PATH>/bin <ISSM_PATH>/lib
\end{lstlisting}
again, where \lstinlinebg|<ISSM_PATH>| is the path to ISSM on disk. Alternatively, you can use the \href{https://www.mathworks.com/help/matlab/matlab_env/what-is-the-matlab-search-path.html}{`Set Path' dialog} to add these paths manually. If you installed ISSM by downloading one of our precompiled distributables, you will also need to run,
\begin{lstlisting}
addpath <ISSM_PATH>/share
\end{lstlisting}
Yet another alternative is to feel the \lstinlinebg|addpath| command as a statement to the \lstinlinebg|matlab| startup command, for example with,
\begin{lstlisting}
matlab -r 'addpath ${ISSM_DIR}/bin ${ISSM_DIR}/lib'
\end{lstlisting}

You can verify that ISSM is findable by running,
\begin{lstlisting}
issmversion
\end{lstlisting}
which should print a message similar to,
\begin{lstlisting}
Ice-sheet and Sea-level System Model (ISSM) Version 4.22
(website: http://issm.jpl.nasa.gov contact:issm@jpl.nasa.gov)

Build date: Wed Sep 18 14:00:06 PDT 2023
Copyright (c) 2009-2023 California Institute of Technology

    to get started type: issmdoc
\end{lstlisting}

To avoid having to manually load ISSM each time you start up MATLAB, you might create an alias in your shell configuration file like,
\begin{lstlisting}
alias matlab-issm="matlab -r 'addpath ${ISSM_DIR}/bin ${ISSM_DIR}/lib'"
\end{lstlisting}
If you will not be using MATLAB's GUI, you might create aliases like,
\begin{lstlisting}
alias matlab_no_gui="matlab -nosplash -nodesktop -nodisplay"
alias matlab-no_gui-issm="matlab -r 'addpath ${ISSM_DIR}/bin ${ISSM_DIR}/lib'"
\end{lstlisting}

\vspace{1cm}\hrule

\subsubsection{Python}
Before starting Python, activate the virtual environment that you set up during installation with, for example,
\begin{lstlisting}
source ${HOME}/.venv/ISSM/bin/activate
\end{lstlisting}

After starting Python, you must tell it where to find ISSM's binaries and libraries by running,
\begin{lstlisting}
import os
import sys
ISSM_DIR = os.getenv('ISSM_DIR')
sys.path.append(ISSM_DIR + '/bin')
sys.path.append(ISSM_DIR + '/lib')
\end{lstlisting}
If you installed ISSM by downloading one of our precompiled distributables, you will also need to run,
\begin{lstlisting}
sys.path.append(os.getenv('ISSM_DIR') + '/share')
\end{lstlisting}

You can verify that ISSM is findable by running,
\begin{lstlisting}
from issmversion import issmversion
\end{lstlisting}
which should print a message similar to,
\begin{lstlisting}
Ice-sheet and Sea-level System Model (ISSM) Version 4.22
(website: http://issm.jpl.nasa.gov contact:issm@jpl.nasa.gov)

Build date: Wed Sep 18 14:00:06 PDT 2023
Copyright (c) 2009-2023 California Institute of Technology
\end{lstlisting}

To avoid having to manually load ISSM each time you start up Python, you might create a startup script, for example, \lstinlinebg|${HOME}/ISSM/bin/sitecustomize.py| with the following contents,
\begin{lstlisting}
#!/usr/bin/env python3
import os
import sys
ISSM_DIR = os.getenv('ISSM_DIR')
sys.path.append(ISSM_DIR + '/bin')
sys.path.append(ISSM_DIR + '/lib')
\end{lstlisting}
Then, add the following to the bottom of your virtual environment activation script (again, for example, \lstinlinebg|${HOME}/.venv/ISSM/bin/activate|):
\begin{lstlisting}
export PYTHONPATH="${HOME}/ISSM/bin:${PYTHONPATH}"
\end{lstlisting}
ISSM should now be findable by Python whether you are in an interactive or non-interactive session as long as you have first activated your virtual environment.

\vspace{1cm}\hrule

\subsubsection{Development Build}
If you configured and compiled ISSM for development, you will have to load ISSM a bit differently.

\emph{MATLAB}
\begin{lstlisting}
addpath <ISSM_PATH>/src/m/dev
devpath
\end{lstlisting}
again, where \lstinlinebg|<ISSM_PATH>| is the path to ISSM on disk.

\emph{Python}
Before starting Python, run,
\begin{lstlisting}
export PYTHONPATH="${ISSM_DIR}/src/m/dev"
export PYTHONSTARTUP="${PYTHONPATH}/devpath.py"
\end{lstlisting}

For IPython users, instead launch the interface with,
\begin{lstlisting}
ipython -i ${ISSM_DIR}/src/m/dev/devpath.py
\end{lstlisting}

\clearpage % Make sure all figures are placed before next section
